---
knit: "bookdown::preview_chapter"
---

```{r, echo=FALSE}
rm(list=ls())
```

# Randomization

Here is a review of existing methods.

## Understanding the basics

Let us write a very simple data-generating process (DGP) for a trivial RCT.

```{r chunk-rct11, warning = FALSE, message = FALSE}

library(mvtnorm)
set.seed(2001)
beta <- 0.5
sigma <- matrix(c(1.0,0.0,
                  0.0,1.0),ncol=2)

## Generate residuals and variables
N <- 20
m <- rmvnorm(N,mean=c(0,0),sigma=sigma)

## Write down the DGP:
Dstar <- m[,1]
D <- ifelse(Dstar>0,1,0)
Y <- beta*D + m[,2]
data <- data.frame(cbind(D,Y))
head(data)
```

The standard way of doing the impact evaluation is then simply a comparison of means, which boils down to an OLS regression.

```{r chunk-rct12, warning = FALSE, message = FALSE}
summary(lm(Y ~ D))
```

@Heckman96


## The randomista debate

@Deaton2010 
@Imbens10 
@Barrett2010 
@Ravallion2012 
@Deaton2016
@Ravallion2018

## Bias and randomization inference

@Young2016a

The basic idea behind randomization inference is to do all of the permutations of treatment status `D`.
We can do this with the `mosaic` package.

```{r chunk-rct13, warning = FALSE, message = FALSE}

# How many unique permutations are there of treatment?
2^20

library(mosaic)
obsdiff = with(data, mean(Y[D==1]) - mean(Y[D==0]))
obsdiff
numsim = 1000
res = do(numsim) * lm(Y ~ shuffle(D), data=data)
pvalue = sum(abs(res$D) > abs(obsdiff)) / numsim
histogram(~ res$D, group = abs(res$D) > abs(obsdiff), 
          n=20, density=FALSE, data=res, xlab="difference between groups",
          main=paste("Permutation test: p=", pvalue))
```

Now do it properly using the `coin` package

```{r chunk-rct14}
library(coin)
oneway_test(Y ~ as.factor(D), alternative = "two.sided", data=data, distribution = "exact")
```

You should also read @Eble2017 and @Athey2016

## Placebo effects and (surprise) homo oeconomicus is alive and kicking

@Malani2006
@Bulte2014
@Chassang2012
@Wing2016
