---
knit: "bookdown::preview_chapter"
---

```{r, echo=FALSE}
rm(list=ls())
```

# Heterogeneity in various flavors

Heterogeneity appears in various forms in impact evaluation.  But it is pretty safe to say that there are two main varieties: (i) heterogeneity which depends on the LHS variable and (ii) that which depends on RHS variables.  

## Quantile regressions and random coefficient models

@Koenker01
@Beck2007

## Essential heterogeneity

@Heckman99a
@Heckman06
@Heckman03
@Ravallion2015

We begin by illustrating what happens when a model is affected by a small amount of essential heterogeneity.

```{r , warning = FALSE, message = FALSE}
# Model with very little essential heterogeneity

set.seed(14381)
N <- 1000
alfa <- 0.20
gamma <- 0.67
epsilon <- rnorm(N,0,1)
sigv = -1.000
sig1 <- 0.12
sig0 <- 0.18
U1 <- sig1*epsilon
U0 <- sig0*epsilon
V <- sigv*epsilon
UD <- pnorm(V/(sigv))
#hist(V)
#hist(UD)
Z <- rnorm(N,-0.026,1.700)
Dstar <- Z - V
D <- ifelse(Dstar>0,1,0)
Y1 <- gamma + alfa + U1
Y0 <- gamma        + U0
Y <- D*Y1 + (1-D)*Y0
# summary(lm(Y ~ D))
# deltaY <- Y1 - Y0
# plot(UD,deltaY)

# Estimate the discrete choice model in
# order to construct the propensity score P(Z)

D.probit <- glm(D ~ Z, family=binomial(link="probit"))
summary(D.probit)
propensity.score <- predict(D.probit, type="response")
summary(propensity.score)

# Consider the propensity score separately
# for treated and untreated individuals

D1 <- ifelse(D == 1, 1, NA)
D0 <- ifelse(D == 0, 1, NA)
propensity.score_1 <- na.omit(propensity.score*D1)
propensity.score_0 <- na.omit(propensity.score*D0)
summary(propensity.score_1)
summary(propensity.score_0)

# Check for range of common support

CS_min <- max(min(propensity.score_1),min(propensity.score_0))
CS_max <- min(max(propensity.score_1),max(propensity.score_0))
CS <- cbind(CS_min,CS_max)
CS_dummy_min <- ifelse(CS_min <= propensity.score,1,NA)
CS_dummy_max <- ifelse(propensity.score <= CS_max,1,NA)
CS_dummy <- CS_dummy_min*CS_dummy_max
summary(CS_dummy)
CS

# Construct histograms of propensity score
# for treated and untreated individuals

propensity.score.hist_1 <- hist(propensity.score_1, br=50)
propensity.score.hist_0 <- hist(propensity.score_0, br=50)
plot(propensity.score.hist_1,main="Histograms of the distributions 
     of the propensity score, for D=1 and D=0",xlab="Propensity score P(Z)", col=adjustcolor("blue", 0.3))
lines(propensity.score.hist_0, lty=2, col=adjustcolor("red", 0.3))

ps <- propensity.score
heckman <- as.data.frame(cbind(Y,D,Z,ps))
attach(heckman)

# Compute simple linear IV estimate

library(AER)
lineariv <- ivreg(Y ~ D | ps, data=heckman)
summary(lineariv)

# Estimate the linear treatment model with the estimated propensity score P(Z)
# Include quadratic and cubic terms in order to test for the presence of 
# essential heterogeneity

# Test the joint significance of the quadratic and cubic terms

ps_2 <- ps^2
ps_3 <- ps^3
Y.ps1 <- lm(Y ~ ps + ps_2 + ps_3, data = heckman)
summary(Y.ps1)
library(car)
linearHypothesis(Y.ps1, c("ps_2 = 0", "ps_3 = 0"))

# Calculate the MTE parametrically

library(margins)
Y.ps2 <- lm(Y ~ ps + I(ps^2) + I(ps^3), data = heckman)
cplot(Y.ps2, "ps", what = "effect", main = "Parametric MTE")

# Nonparametric specification

library(np)

# Use a simple local linear regression with 
# bootstrapped or asymptotic standard errors

bw0 <- npregbw(xdat=ps, ydat=Y, regtype="ll", bwmethod="cv.aic")
Y.np0 <- npreg(bws = bw0, gradient=TRUE)
summary(Y.np0)
plot(Y.np0, plot.errors.method="bootstrap")
plot(Y.np0, plot.errors.method="bootstrap", gradient=TRUE, 
     ylab="MTE", xlab="Propensity score")
# The MTE is valid over the region of common support
mte <- Y.np0$grad*CS_dummy
ATE <- mean(mte, na.rm=TRUE)
abline(h=ATE)
mte_se <- Y.np0$gerr*CS_dummy
ATE_se <- mean(mte_se, na.rm=TRUE)

# Compute the treatment weights

uD <- qnorm(ps)
P.probit <- glm(D ~ uD, family=binomial(link="probit"))
summary(P.probit)
omega_TT_n <- 1 - predict(P.probit, type="response")
omega_TT_d <- sum(omega_TT_n)
omega_TT <- omega_TT_n / omega_TT_d
omega_TUT_n <- predict(P.probit, type="response")
omega_TUT_d <- sum(omega_TUT_n)
omega_TUT <- omega_TUT_n / omega_TUT_d
check <- sum(omega_TT)
print(check)

TT <- sum(omega_TT*mte, na.rm=TRUE)
TT_se <- sum(omega_TT*mte_se, na.rm=TRUE)
TUT <- sum(omega_TUT*mte, na.rm=TRUE)
TUT_se <- sum(omega_TUT*mte_se, na.rm=TRUE)
parameters <- cbind("", "ATE", "TT", "TUT")
mean.effects <- cbind("Parameter", round(ATE,digits=3),round(TT,digits=3),round(TUT,digits=3))
mean.effects_se <- cbind("s.e", round(ATE_se,digits=3), round(TT_se,digits=3), round(TUT_se,digits=3))
print("Region of common support")
print(CS)
effects <- rbind(parameters, mean.effects, mean.effects_se)
print(effects)

plot(Y.np0, plot.errors.method="asymptotic", gradient=TRUE, 
     ylab="MTE", xlab="Propensity score")
abline(h=ATE, lty=1)
abline(h=TT, lty=1, col="blue")
abline(h=TUT, lty=1, col="red")
```


```{r , warning = FALSE, message = FALSE}
# MODEL WITH ESSENTIAL HETEROGENEITY

set.seed(14381)
N <- 1000
alfa <- 0.20
gamma <- 0.67
epsilon <- rnorm(N,0,1)
sigv = -1.000
sig1 <- 0.12
sig0 <- -0.50
U1 <- sig1*epsilon
U0 <- sig0*epsilon
V <- sigv*epsilon
# UD <- pnorm(V/(sigv))
# hist(UD)
Z <- rnorm(N,-0.026,1.700)
Dstar <- Z - V
D <- ifelse(Dstar>0,1,0)
Y1 <- gamma + alfa + U1
Y0 <- gamma        + U0
Y <- D*Y1 + (1-D)*Y0
# summary(lm(Y ~ D))
# deltaY <- Y1 - Y0
# plot(UD,deltaY)
# 

# Estimate the discrete choice model in
# order to construct the propensity score P(Z)

D.probit <- glm(D ~ Z, family=binomial(link="probit"))
summary(D.probit)
propensity.score <- predict(D.probit, type="response")
summary(propensity.score)

# Consider the propensity score separately
# for treated and untreated individuals

D1 <- ifelse(D == 1, 1, NA)
D0 <- ifelse(D == 0, 1, NA)
propensity.score_1 <- na.omit(propensity.score*D1)
propensity.score_0 <- na.omit(propensity.score*D0)
summary(propensity.score_1)
summary(propensity.score_0)

# Check for range of common support

CS_min <- max(min(propensity.score_1),min(propensity.score_0))
CS_max <- min(max(propensity.score_1),max(propensity.score_0))
CS <- cbind(CS_min,CS_max)
CS_dummy_min <- ifelse(CS_min <= propensity.score,1,NA)
CS_dummy_max <- ifelse(propensity.score <= CS_max,1,NA)
CS_dummy <- CS_dummy_min*CS_dummy_max
summary(CS_dummy)
CS

# Construct histograms of propensity score
# for treated and untreated individuals

propensity.score.hist_1 <- hist(propensity.score_1, br=50)
propensity.score.hist_0 <- hist(propensity.score_0, br=50)
plot(propensity.score.hist_1,main="Histograms of the distributions 
of the propensity score, for D=1 and D=0",xlab="Propensity score P(Z)", col=adjustcolor("blue", 0.3))
lines(propensity.score.hist_0, lty=2, col=adjustcolor("red", 0.3))

ps <- propensity.score
heckman <- as.data.frame(cbind(Y,D,Z,ps))
attach(heckman)

# Compute simple linear IV estimate

library(AER)
lineariv <- ivreg(Y ~ D | ps, data=heckman)
summary(lineariv)

# Estimate the linear treatment model with the estimated propensity score P(Z)
# Include quadratic and cubic terms in order to test for the presence of 
# essential heterogeneity

# Test the joint significance of the quadratic and cubic terms

ps_2 <- ps^2
ps_3 <- ps^3
Y.ps1 <- lm(Y ~ ps + ps_2 + ps_3, data = heckman)
summary(Y.ps1)
library(car)
linearHypothesis(Y.ps1, c("ps_2 = 0", "ps_3 = 0"))

# Calculate the MTE parametrically

library(margins)
Y.ps2 <- lm(Y ~ ps + I(ps^2) + I(ps^3), data = heckman)
cplot(Y.ps2, "ps", what = "effect", main = "Parametric MTE")

# Nonparametric specification

library(np)

# Use a simple local linear regression with asymptotic standard errors

bw0 <- npregbw(xdat=ps, ydat=Y, regtype="ll", bwmethod="cv.aic")
Y.np0 <- npreg(bws = bw0, gradient=TRUE)
summary(Y.np0)
plot(Y.np0, plot.errors.method="asymptotic")
plot(Y.np0, plot.errors.method="asymptotic", gradient=TRUE, 
     ylab="MTE", xlab="Propensity score")
# The MTE is valid over the region of common support
mte <- Y.np0$grad*CS_dummy
ATE <- mean(mte, na.rm=TRUE)
abline(h=ATE)
mte_se <- Y.np0$gerr*CS_dummy
ATE_se <- mean(mte_se, na.rm=TRUE)

# Compute the treatment weights

uD <- qnorm(ps)
P.probit <- glm(D ~ uD, family=binomial(link="probit"))
summary(P.probit)
omega_TT_n <- 1 - predict(P.probit, type="response")
omega_TT_d <- sum(omega_TT_n)
omega_TT <- omega_TT_n / omega_TT_d
omega_TUT_n <- predict(P.probit, type="response")
omega_TUT_d <- sum(omega_TUT_n)
omega_TUT <- omega_TUT_n / omega_TUT_d
check <- sum(omega_TT)
print(check)

TT <- sum(omega_TT*mte, na.rm=TRUE)
TT_se <- sum(omega_TT*mte_se, na.rm=TRUE)
TUT <- sum(omega_TUT*mte, na.rm=TRUE)
TUT_se <- sum(omega_TUT*mte_se, na.rm=TRUE)
parameters <- cbind("", "ATE", "TT", "TUT")
mean.effects <- cbind("Parameter", round(ATE,digits=3),round(TT,digits=3),round(TUT,digits=3))
mean.effects_se <- cbind("s.e", round(ATE_se,digits=3), round(TT_se,digits=3), round(TUT_se,digits=3))
print("Region of common support")
print(CS)
effects <- rbind(parameters, mean.effects, mean.effects_se)
print(effects)

plot(Y.np0, plot.errors.method="asymptotic", gradient=TRUE, 
     ylab="MTE", xlab="Propensity score")
abline(h=ATE, lty=1)
abline(h=TT, lty=1, col="blue")
abline(h=TUT, lty=1, col="red")

# Can also use a more sophisticated Li and Racine bandwidth estimation
# with bootstrapped standard errors
# In this case, it does not improve performance
# bw1 <- npregbw(formula = Y ~ ps)
# Y.np1 <- npreg(bws = bw1, gradient=TRUE)
# summary(Y.np1)
# plot(Y.np1, plot.errors.method="bootstrap")
# plot(Y.np1, gradients=TRUE, plot.errors.method="bootstrap",
#      ylab="MTE", xlab="Propensity score")
# mte <- Y.np1$grad*CS_dummy
# ATE <- mean(mte, na.rm=TRUE)
# abline(h=ATE)
# mte_se <- Y.np1$gerr*CS_dummy
# ATE_se <- mean(mte_se, na.rm=TRUE)

sessionInfo()

```

# References {-}
